---
title: "Tutorial Part I - Quick introduction to R"
author: "Stephanie Kramer-Schadt, Alex Courtiol, Moritz Wenzler"
date: "`r Sys.Date()`"
output:
  #prettydoc::html_pretty:
  #  theme: cayman
  html_document:
    code_folding: show
    link-citations: yes
    toc: yes
    toc_depth: 4
    toc_float: yes
  slidy_presentation: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = 'C:/Users/kramer/PopDynIZW Dropbox/Steph Kramer')
```

# Basics

The first outline of this course has been developed in 2012 by Alex Courtiol and has ever since 'evolved' in courses I gave and polished by Moritz. This short intro contains the most important commands and functions to handle and manipulate data and to follow the course 'biodiversity dynamics. Open your RStudio and type or paste the lines in the grey boxes into your R script, run the lines and follow the course... Or: Click on *Code/Download ZIP* and unzip the folder. However, for learning reasons, I actually recommend to even type the functions and commands by hand.

For deepening your R skills, I recommend the tutorials of  

* https://ourcodingclub.github.io/tutorials.html

They provide screenshots and glossaries of the most important functions in the end.

## Setting the work directory

Please set your OWN working directory!
Function **paste** 

```{r}
getwd() #check where you are on your PC, change it to your folder
# you can skip the following if you are already working in a project
setwd('C:/Users/kramer/PopDynIZW Dropbox/Steph Kramer/_GitHub/Course1_IntroR')
#-- no need to do so if you created an R-project --

# to keep folders clean, store all output in a separate folder
output_wd <- paste("C:/Users/kramer/PopDynIZW Dropbox/Steph Kramer", "/", "A_Borrar", sep = '')
```

## Basics: R for calculus
hashtag # for comments

```{r}

# "comment"
1+1
# store calculus etc in an object
what_is_one_plus_one <- 1+1
# display result
what_is_one_plus_one
# assign character
what_is_one_plus_one <- 'two'
```

## Basics: Common mistakes

### Spelling - Case sensitivity

```{r}
# Find the errors - try these lines
#What_is_one_plus_one
#what_is_one_plusone
#what_is_one_plus_two
```


### Wrong assignment

```{r, eval = F}
 1 + what_is_one_plus_one
# why does this not work?

 1 + what_is_one_plus_one <- 1 + 2
# nor this?
```

### Correct assignment

```{r}
what_is_one_plus_one <- 1 + 1
1 + what_is_one_plus_one
```

## Basics: Getting help

```{r, eval=FALSE}
# <function_name> or <function_name>() shows the function
citation() # how to cite R
?citation() # getting help use ? (or??) or
help(citation)
```

## Basics: Functions 

functions indicated by ()

```{r, eval = F}
mean() # what's wrong?
```

this function needs further parameters (called Arguments in the help)
```{r}
?mean() # which parameters are needed?
```



### Assign values to the function

```{r}
#- Use the command 'c' - it combines values into a vector or list
vector_of_numbers <- c(22, 65, 101, 86, 110)
# e.g. wingspan of dragonflies in mm 
vector_of_numbers
mean(x = vector_of_numbers)
```

### Arguments

```{r}
mean(x = vector_of_numbers) #n.b. better use argument names (x= ...)
```

```{r, eval = F}
mean(y = vector_of_numbers) #what's wrong?
```

```{r}
# analogously: more useful functions/ commands:
sqrt(x = vector_of_numbers) # exp(), sin(), log()...
sum(x = vector_of_numbers)
```


### Applying functions

```{r}
# Assign the assigned values
wsp_drfly <- vector_of_numbers #wingspan of dragonflies
wsp_drfly
mean(x = wsp_drfly)

# Adding data, e.g. new measurements
wsp_drfly2 <- c(vector_of_numbers, wsp_drfly ,69 ,73)
wsp_drfly2
```

## Useful functions

```{r}
table(wsp_drfly2) # returns the counts
sort(x = wsp_drfly2)
min(wsp_drfly2) #analogous max, var, sd, range, median
```


```{r}
unique(x = wsp_drfly2) # n.b. - does not sort vector
length(x = wsp_drfly2)
summary(object = wsp_drfly2)
```

## Operators 

### Arithmetic operators

```{r}
?Arithmetic
wsp_drfly2 * 2
wsp_drfly2 * 10 - 100
```

### Logical operators

```{r}

?base::Logic 
wsp_drfly2 > 10 # >=, <, <=, &, | , !=, ==
wsp_drfly2 == 75 # n.b. difference to = or <- !!!!
```

<br><br>

# Data Types and Data Manipulation 

## Create Data Set with Vectors

#### Repeat data with command *rep*

```{r}

# wingspan of butterflies in mm 
wsp_butfly2 <- c(12, 48, 79, 135, 189, 88, 15, 63, 21, 81, 148, 50)
data_length <- length(x = wsp_butfly2) #12
wsp <- c(wsp_drfly2, wsp_butfly2)
insect_type <- factor(x = c(rep('dragonfly', data_length),
                            rep('butterfly', data_length))
                      )
# n.b. correct position of parentheses when functions nested
insect_type
```

## Dataframe [df]

```{r, tidy=TRUE}
wsp_table <- data.frame(wingspan = wsp, insect = insect_type)
head(x = wsp_table) # head() returns first 6 rows # tail()
wsp_table

## Retrieving data from df
str(object = wsp_table)
wsp
wsp_table$wingspan # access data with dollar sign
#**What is the average wsp?**
```

### Indexing elements

#### Indexing a vector[element]

```{r}
vector_of_numbers[3]
# vector_of_numbers[c(1, 3:5)]
# why c(...)? Because the comma is another dimension.
# dimension in R: [row, column]
# Not run: vector_of_numbers[1, 3:5] #error
```

#### Indexing with [row,column]

```{r}
wsp_table[2, 2] # == wsp_table$insect[2] #!
wsp_table[2, 1] # == wsp_table$wingspan[2] #!
wsp_table[2, ]
```

```{r}
wsp_table[, 2]
```

```{r, eval = F}
 wsp_table[, 3] #what's wrong?
```

```{r}
wsp_table[c(2:5), 2]
```

### Dataframe queries

#### Function *tapply* / table apply 

**What is the average wingspan PER insect species?**

```{r}
mean(x = wsp_table$wingspan[wsp_table$insect == 'dragonfly']) # == !
#- more elegant: function 'tapply'
tapply(X = wsp_table$wingspan, INDEX = wsp_table$insect, FUN = mean)
```

#### Subsetting 

##### simple: function *subset*

```{r}
wsp_sub <- subset(x = wsp_table, subset = wsp_table$wingspan <= 65)
nrow(wsp_sub) # try also: dim(wsp_sub),dim(wsp_sub)[1]
wsp_sub
```

##### A bit more complicated: function *which*

```{r}
wsp_tmp <- which(x = wsp_table$wingspan <= 65)
wsp_tmp
wsp_which <- wsp_table[wsp_tmp,]
wsp_which
```

### Data Manipulation

#### Removing rows/columns 

Omit rows using a minus sign

```{r}
x <- wsp_which
x
x2 <- x[-c(1, 3), ] #remove rows
x2
```


Omit columns

```{r}
x22 <- x$wingspan #removes (actually keeps) columns
x22
```

#### Adding rows/columns I

##### Combine data with *rbind()*

```{r}
x3 <- rbind(x2, wsp_table)
x3[1:10,] # n.b. new row names 41, 91 !
```

##### Combine data with *cbind()*

```{r}
x4 <- cbind(vector_of_numbers[1:3], vector_of_numbers[3:5])
x4
colnames(x4) <- c('col1', 'col2')
rownames(x4) <- c('A', 'B', 'C')
x4
```

## List

#### list per grouping factor (here: insect type)

```{r}
wsp_list <- list(dragonfly = wsp_drfly2, butterfly = wsp_butfly2)
str(wsp_list)
wsp_list$dragonfly # == wsp_list[['dragonfly']]
lapply(X = wsp_list, FUN = mean) # OR: mean(wsp_list$dragonfly)
```

## Matrix 

Definition per row or col, looses headers!
matrix setup per row

```{r}
wsp_matrix <- matrix(data = c(wsp_drfly2, wsp_butfly2), nrow = 2)
wsp_matrix
# Does this format ring a bell????
rownames(x = wsp_matrix) <- c('dragonfly', 'butterfly')
wsp_matrix[, 1]
wsp_matrix['dragonfly', ] # == wsp_matrix[1, ]
```

matrix setup per column:

```{r}
wsp_matrix <- matrix(data = c(wsp_drfly2, wsp_butfly2), ncol = 2)
wsp_matrix
#head(wsp_matrix)
#str(wsp_matrix)
wsp_matrix[, 1]
colnames(x = wsp_matrix) <- c('dragonfly', 'butterfly')
#wsp_matrix$dragonfly
wsp_matrix[, 'dragonfly'] # == wsp_matrix[, 1]
```

#### Extracting data from matrix 

```{r}
#wsp_matrix <- matrix(c(wsp_drfly2,wsp_butfly2), ncol=2)
#colnames(wsp_matrix) <- c('dragonfly','butterfly')

mean(x = wsp_matrix[, 1])
apply(X = wsp_matrix, MARGIN = 2, FUN = mean)
#check command, what is meant with 'by 2'?
#check command, what happens if you use 1 instead?
```

## Missing Data NA

```{r}
wsp_fly2 <- c(1,5,4,11,8,2,NA,4,2,9,8,NA)
mean(x = wsp_fly2)
# type ?mean  # How to deal with missing data?

mean(x = wsp_fly2, na.rm=T)
```

```{r}
wsp_l <- list(dragonfly = wsp_drfly2, butterfly = wsp_butfly2, fly = wsp_fly2)
lapply(X = wsp_l, FUN = mean)
lapply(X = wsp_l, FUN = mean, na.rm = T)
```

## Data Manipulation: Summary

```{r}
#TRY
# class(wsp_table) #class(wsp_matrix) #class(wsp_list)
# on vectors: #class(insect_type) # class(wsp_l$dragonfly)
# on skalars: #class(what_is_one_plus_one)
```

For effective data manipulation, learn 'tidyverse','dplyr' and 'broom' piping language.

<br><br>

# Managing Data 

## Saving and Reading Files 

```{r}
getwd() # to change use setwd(), e.g. setwd('../Output')
write.csv(x = wsp_table, file = "wsp.csv", row.names = FALSE)
wsp <- read.csv(file = "wsp.csv", header = T)
head(x = wsp)
```

<br><br>

# Plotting with R 

## Plotting: Basics

**plot()** is the universal function for scatter plots or line diagrams

```{r}
# this does not work because you need to specify numeric values, 
# column 'insect' is a character
# plot(x = wsp)

# solution
plot(x = wsp$wingspan)

```


Use command *ifelse* to define colors PER species

```{r}
color_for_insect_type <- ifelse(test = wsp$insect == 'dragonfly', yes = 'red', no = 'blue')
# color_for_insect_type
plot(x = wsp$wingspan, col= color_for_insect_type)
```


**par()** to pimp up your plot

```{r}
# now have a look at
?base::plot #and
?par # of how to make the plots more pretty and how to set graphical parameters

plot(x = wsp$wingspan, col = color_for_insect_type, pch = 3,
      xlab = "unsorted values", ylab = "wsp [mm]", cex = 3,
      cex.lab = 1.2, cex.axis = 0.5, las = 2,
      col.axis = "blue")
```

## Plotting: Practical thingies 
### plot-function
#### Abline - Drawing lines (h=horizontal, v= vertical, or "intercept, slope")

```{r}
?abline
plot(x = wsp$wingspan, col = color_for_insect_type)
abline(v = 12.5, lty = "dashed", col = "red")
                             #lty = line type
abline(h = 75, lty = "dotted")
```


#### Drawing *arrows* and adding *text*

```{r}
plot(x = wsp$wingspan, col = color_for_insect_type,pch = 16, cex = 3)
arrows(x0 = 5, y0 = 125, x1 = 5, y1 = 82, col = "purple")
arrows(x0 = 15, y0 =  70, x1 = 17, y1 =  75, col = "black", lwd = 3) #lwd=line width
text(x = 3, y = 70, labels = "DRAGONFLY",cex=0.8)
```

## Other plot types

```{r}
#First, create new data set called " mydata" with new column "wind_speed":
wind_speed <- rep(x = 50:70, length.out = nrow(wsp))
mydata <- cbind(wsp, wind_speed)
head(x = mydata)
```

### *boxplot*, histogram (*hist*)

```{r}
par(mfrow = c(1, 2))
boxplot(formula = mydata$wingspan ~ mydata$wind_speed)
hist(x = mydata$wingspan, main = '')
```

### *hist* continued - the role of breaks and freq = F

```{r}
par(mfrow = c(1,2))
hist(x = mydata$wingspan, breaks = 2, main = '')
hist(x = mydata$wingspan, 
     breaks = seq(min(mydata$wingspan),
                  max(mydata$wingspan), 
                  by = 3), 
     freq = F, col = 'lightgrey', main = '')
```

It is also possible to save the results of the histogram in an object
```{r}
my_h <- hist(x = mydata$wingspan, 
             breaks = seq(min(mydata$wingspan), 
                          max(mydata$wingspan), by = 3), plot = FALSE)
my_h
```


### barplot

```{r}
my_bp <- table(mydata$wingspan)
my_bp # note the difference to histograms ($counts)!
barplot(my_bp) # WHY is there a maximum of 3 counts in the histogram and 2 in the barplot?
```


### Conditional plot *coplot*

```{r}
coplot(formula = mydata$wingspan ~ mydata$wind_speed | mydata$insect, pch = 16)
```

## Plotting: Margins and multiple figures

### Graphics device division

```{r}
#mar = margin of figure in blue, oma = outer margin in green
par(mfrow = c(1, 2), mar = c(5, 4, 1, 1), oma = c(1.5, 2, 1, 1))
plot(x = mydata$wingspan, col = color_for_insect_type, pch = 16, cex = 3)
boxplot(formula = mydata$wingspan ~ mydata$insect)
box(which = 'plot', col = 'red')
box(which = 'figure', col = 'blue', lty = 'dashed')
box(which = 'outer', col = 'green' , lwd = 5)
```

## Saving Plots

works for jpg, pdf,png, wmf and postscript

```{r}
#- Single Plot
pdf(file = 'MyPdf.pdf', width =  10, height = 10)
with(data = mydata, expr = {coplot(wingspan ~ wind_speed | insect, pch=16)})
dev.off()
#- Multiple Plot
jpeg(file = 'MyJPG.jpg') #n.b.! Correct output folder path!, e.g.:
# jpeg(file =  paste0(output_wd,'/MyJPG.jpg')) #don't forget the /
par(mfrow = c(1, 2))
barplot(my_bp, col = rainbow(17), main = 'Barplot', xlab = 'wingspan', ylab = 'Count')
hist(x = mydata$wingspan, 
     breaks = seq(min(mydata$wingspan), 
                  max(mydata$wingspan), 
                  by = 3), 
     freq = T, 
     col = rainbow(6))
dev.off()

```

## Fancy plotting with ggplot2

Let's be honest, the above plots are ugly. For making nice plots you can use in 
theses or publications, change to ggplot.
<br>

I strongly recommend the course by Cedric Scherer:  
* https://www.cedricscherer.com/2019/08/05/a-ggplot2-tutorial-for-beautiful-plotting-in-r/
<br>
or by the coding club:
<br>
* https://ourcodingclub.github.io/tutorials/trends-analysis-dataviz/
<br>

The *ggplot* language works in layers to be built and added. First, load the library,
and plot the canvas for 'mydata':
```{r}
library(ggplot2) # don't forget to install the package
ggplot(data=mydata) # nothing happens but a grey panel appearing
```

Now, add layers with the '+' sign. First, specify the type of plot (geom_point),
and then which column to plot and how they should look like with 'aes', 
standing for 'aesthetics':
```{r}
ggplot(data=mydata) + geom_point(aes(x=wind_speed, y =wingspan))
```

Pimp up the plot, look at the specifications e.g. here:
https://ggplot2.tidyverse.org/articles/ggplot2-specs.html
```{r}
ggplot(data=mydata) + geom_point(
                                 aes(x=wind_speed, y =wingspan),
                                 shape=17,
                                 size=5,
                                 col='grey70',
                                 alpha = 0.5
                                 )
```

If you like the look, you can even store this plot in an object
```{r}
myggplot1 <- ggplot(data=mydata) + 
                geom_point(aes(x=wind_speed, y =wingspan),
                           shape=17,
                           size=5,
                           col="red",
                           alpha = 0.4) #makes shapes transparent
#to plot, call the object
myggplot1
```

Add the labels, and your favourite plotting theme:  
https://ggplot2.tidyverse.org/reference/ggtheme.html
or check 'library(ggthemes)'
```{r}
myggplot1 + 
  theme_dark() + 
  labs(x= 'Wind Speed', y= 'Wing span [mm]', title='',size= 15)
```

Now make the plot per insect with 'facet':

```{r}
myggplot1 + 
  theme_light() + 
  labs(x= 'Wind Speed', y= 'Wing span [mm]', title='',size= 15) +
  facet_grid(. ~ insect)
```

Now, make a nice boxplot adding the points with *geom_jitter*

```{r}
ggplot(mydata, aes(insect, wingspan)) +
  geom_boxplot() +
  geom_jitter(color = wind_speed,width = .1, alpha = .5,size=5) + 
  labs(x = "Insect taxon", y = "Wing span [mm]")
```

An informative alternative are violin-plots:

```{r}
ggplot(mydata, aes(insect, wingspan)) +
  geom_violin(fill = "gray80", size = 1, alpha = .5) +
  geom_jitter(color = 'grey50',width = .1, alpha = .5, size=10) + 
  labs(x = "Insect taxon", y = "Wing span [mm]")
```

Or another beautiful possibility, the ridge plot. Please note: this is a density plot,
forcing beginning and end through zero. But: we never measured wing spans of zero!

```{r}
library(ggridges) #install package
ggplot(mydata, aes(wingspan, factor(insect))) +
   geom_density_ridges(fill = "gray90") +
   labs(x = "Wing span [mm]", y = "Taxon") +
   theme_light()
```


The following is playing around with images  

```{r}
library(rphylopic) # install first
library(ggpubr)    # for arranging the plots

#import phylopic (go to phylopic.org -> browse image -> click on image to get url)
drfly <- image_data("2894776f-a2cf-4475-b13d-6d8ea1607ea4", size = 128)[[1]]
bufly <- image_data("ef0f4ab5-9714-45dc-8760-57e9080f941e", size = 128)[[1]]

p1 <- ggplot(data=mydata) + 
             geom_point(aes(x=wind_speed, y =wingspan)) + 
             theme_light() + 
             labs(x= 'Wind Speed', y= 'Wing span [mm]', title='',size= 15) +
             add_phylopic(img=drfly)

###
p2 <- ggplot(data=mydata,aes(x=wind_speed, y =wingspan)) + 
                theme_light() + 
                geom_point(color = rgb(0,0,0,0))

sub_drag <- subset(mydata, insect == 'dragonfly')
sub_butt <- subset(mydata, insect == 'butterfly')

for (i in 1:nrow(sub_drag)) {
  p2 <- p2 + add_phylopic(drfly, 1, 
                           sub_drag$wind_speed[i], 
                           sub_drag$wingspan[i],10,'blue')}
for (j in 1:nrow(sub_butt)) {
  p2 <- p2 + add_phylopic(bufly, 1, 
                           sub_butt$wind_speed[j], 
                           sub_butt$wingspan[j],10,'red')}

###
p3 <- ggplot(mydata, aes(wingspan, factor(insect))) +
                geom_density_ridges(fill = "gray90") +
                labs(x = "Wing span [mm]", y = "Taxon") +
                theme_light() + 
                add_phylopic(img=bufly, alpha=1, 
                             x = 50, y = 1.5, ysize=20) +
                add_phylopic(img=drfly, alpha=1, 
                             x = 80, y = 3, ysize=25) 

# TRY:
# ggarrange(p1, p2, p3,ncol = 2, nrow = 2)

ggarrange(
  p3,                # First row with ridge plot
  # Second row with other plots
  ggarrange(p1, p2, ncol = 2, labels = c("B", "C")), 
  nrow = 2, 
  labels = "A"       # Label of the line plot
  ) # %>% ggexport(filename = "multifigure.png")# and save the plot

```





Happy with the plot? Then save it:

```{r}
myridgeplot <- ggplot(mydata, aes(wingspan, factor(insect))) +
   geom_density_ridges(fill = "gray90") +
   labs(x = "Wing span [mm]", y = "Taxon") +
   theme_light()

myfilename <- "ridges.pdf"

ggsave(myridgeplot, file=myfilename, width=4, height=4)
```

And find more inspiration here: https://www.r-graph-gallery.com/

<br><br>

# Advanced - Programming with R (shortly) 

This is very helpful for understanding your own data (i.e. simulate data where
you know exactly what happens and where you know what should come out) or
when you have repeated operations to do.

## Simulate your own data

R has a lot of distribution functions:  
https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Distributions.html

```{r, tidy=TRUE}
seq_data  <- seq(from = 1, to = 100, by = 1)     # create the x-axis
uni_data  <- runif(n = 100, min = 1, max = 10)   # create random uniform data between 1-10
norm_data <- rnorm(n = 100, mean = 1, sd = 0.5)  # create normally distributed data

#simulate trend = y-axis: intercept with noise + (estimate * variable)
trend_data <-  norm_data + (0.1 * seq_data)

par(mfrow=c(1,1))
plot(x = seq_data, y = uni_data, pch = 15, ylim = c(0,11)) #uniform -> runif
points(x = norm_data, pch = 20, col = 'red') #Gaussian 
points(x = trend_data, pch = 17, col = 'blue') #trend
```

### Just for fun: some stats....

```{r}
my.lm <- lm(formula = trend_data ~ seq_data)
#abline(my.lm)
summary(object = my.lm) #surprise.... the intercept is ~1 and estimate ~ 0.1
```

## The *for* loop

saves a lot of time to automate operations, ie. do the same operation on 1000 datasets

```{r}
#- for repetitive operations
#- *i* is used as variable that is incremented
for(i in 1:4)
{ #begin
  print(x = i)
  if (i == 4) print (x = 'end of lop')
} # end
?'for'
#other useful control flow constructs are: *while*, *repeat*, *if*
```

## Writing your own function

Also for repetitive actions and often used operations. the basic syntax is:  
function(*argument*){1 + 1 - *argument*}  
The function needs an argument, like e.g. a vector with numbers in the case 
of the function mean(), and in the {} some operations are done with the argument.

```{r}
f_logit <- function(a) {-8.13 + (12.55 * a)}
f_prob <- function(x) {1/ (1 + exp(x = -x))} # inverse logit

# ff = e.g. frequency of forest in a cell
ff <- seq(from = 0, to = 1 ,by = 0.05)

the_logit  <- f_logit(a = ff)
pres_prob  <- f_prob(x = the_logit)

#pres_prob
plot(x = ff,y = pres_prob, type = 'l')
```


### Logistic regression and inverse logit

Similar to example above

```{r, tidy=TRUE}
# define functions at the beginning of a script:
f_logit <- function(var_value, intcpt, slp) {intcpt + (slp * var_value)}
f_prob  <- function(x) {1 / (1 + exp(x = -x))} #inverse logit

# simulate trend or load data:
seq_data  <- seq(from = 0, to = 1, by=0.05)
trend_data <- pbinom(q = 1:length(seq_data), size = length(seq_data), prob = 0.7)
plot(x = seq_data, y = trend_data, pch = 15)

# analyse data
my.lm <- glm(formula = trend_data ~ seq_data, family = 'binomial', na.action = na.omit)
summary(my.lm)

# store the output; access my.lm object via: 
#str(my.lm)
myintercept <- my.lm$coefficients[[1]]
myslope     <- my.lm$coefficients[[2]]

# predict it for a certain range of variable values (seq_data)...
mylogit   <- f_logit(var_value = seq_data, intcpt = myintercept, slp = myslope)
pres_prob <- f_prob(x = mylogit)

# strictly linear relationship on the logit scale:
plot(x = seq_data, y = mylogit, type = 'l', lwd = 2)

# on the response / inverse logit scale:
plot(x = seq_data, y = pres_prob, type = 'l', lwd = 2)

# ...or predict it only for a single value:
myvalue <- 0.7
mylogit_s   <- f_logit(var_value = myvalue, intcpt = myintercept, slp = myslope)
(pres_prob_s <- f_prob(x = mylogit_s))

abline(v=myvalue, lty=2,lwd=0.1)
abline(h=pres_prob_s, lty=2,lwd=0.1)
```

<br>
# END



