---
title: "Tutorial Part I - Quick introduction to R"
author: "Stephanie Kramer-Schadt, Alex courtiol, Moritz Wenzler"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    link-citations: yes
    toc: yes
    toc_depth: 4
    toc_float: yes
  slidy_presentation: default
editor_options:
  chunk_output_type: console
---
# Basics

## Setting the work directory

Please set your OWN working directory!
Function **paste** 

```{r}
setwd('c:/Users/kramer/Dropbox/_TU_BiodivDynamics2019/Day1_RIntro')
output_wd <- paste("c:/Users/kramer/Dropbox/_TU_BiodivDynamics2019", "/", "Output", sep = '')
```

## Basics: R for calculus
hashtag # for comments

```{r}

# "comment"
1+1
# store calculus etc in an object
what_is_one_plus_one <- 1+1
# display result
what_is_one_plus_one
# assign character
what_is_one_plus_one <- 'two'
```

## Basics: Common mistakes

### Spelling - Case sensitivity

```{r}
#What_is_one_plus_one
#what_is_one_plusone
#what_is_one_plus_two
```


### Wrong assignment

```{r, eval = F}
 1 + what_is_one_plus_one
# why does this not work?

 1 + what_is_one_plus_one <- 1 + 2
# nor this?
```

### Correct assignment

```{r}
what_is_one_plus_one <- 1 + 1
1 + what_is_one_plus_one
```

## Basics: Getting help

```{r}
# <function_name> or <function_name>() shows the function

 citation() # how to cite R
 ?citation() # getting help use ? (or??) or
 help(citation)
```

## Basics: Functions 

functions indicated by ()

```{r, eval = F}
mean() # what's wrong?
```

this function needs further parameters
```{r}
?mean() # which ones?
```



### Assign values to the function

```{r}
#- Use the command 'c' - it combines values into a vector or list
vector_of_numbers <- c(22, 65, 101, 86, 110)
# e.g. wingspan of dragonflies in mm 
vector_of_numbers
mean(x = vector_of_numbers)
```

### Arguments

```{r}
mean(x = vector_of_numbers) #n.b. better use argument names
```

```{r, eval = F}
mean(y = vector_of_numbers) #what's wrong?
```

```{r}
# analogously: more useful functions/ commands:
sqrt(x = vector_of_numbers) # exp(), sin(), log()...
sum(x = vector_of_numbers)
```


### Applying functions

```{r}
# Assign the assigned values
wsp_drfly <- vector_of_numbers #wingspan of dragonflies
wsp_drfly
mean(x = wsp_drfly)

# Adding data, e.g. new measurements
wsp_drfly2 <- c(vector_of_numbers, wsp_drfly ,69 ,73)
wsp_drfly2
```

## Useful functions

```{r}
table(wsp_drfly2) # returns the counts
sort(x = wsp_drfly2)
min(wsp_drfly2) #analogous max, var, sd, range, median
```


```{r}
unique(x = wsp_drfly2) # n.b. - does not sort vector
length(x = wsp_drfly2)
summary(object = wsp_drfly2)
```

## Operators 
###Arithmetic operators

```{r}

?Arithmetic
wsp_drfly2 * 2
wsp_drfly2 * 10 - 100
```

###Logical operators

```{r}

?Logic
wsp_drfly2 > 10 # >=, <, <=, &, | , !=, ==
wsp_drfly2 == 75 # n.b. difference to = or <- !!!!
```

<br><br>

# Data Types and Data Manipulation 

## Create Data Set with Vectors

#### Repeat data with command *rep*

```{r}

# wingspan of butterflies in mm 
wsp_butfly2 <- c(12, 48, 79, 135, 189, 88, 15, 63, 21, 81, 148, 50)
data_length <- length(x = wsp_butfly2) #12
wsp <- c(wsp_drfly2, wsp_butfly2)
insect_type <- factor(x = c(rep('dragonfly', data_length),
                            rep('butterfly', data_length))
                      )
# n.b. correct position of parentheses when functions nested
insect_type
```

## Dataframe [df]

```{r}
wsp_table <- data.frame(wingspan = wsp, insect = insect_type)
head(x = wsp_table) # head() returns first 6 rows # tail()
wsp_table

## Retrieving data from df
str(object = wsp_table)
wsp
wsp_table$wingspan # access data with dollar sign
#**What is the average wsp?**
```

### Indexing elements

#### Indexing a vector[element]

```{r}
vector_of_numbers[3]
# vector_of_numbers[c(1, 3:5)]
# why c(...)? Because the comma is another dimension.
# dimension in R: [row, column]
# Not run: vector_of_numbers[1, 3:5] #error
```

#### Indexing with [row,column]

```{r}
wsp_table[2, 2] # == wsp_table$insect[2] #!
wsp_table[2, 1] # == wsp_table$wingspan[2] #!
wsp_table[2, ]
```

```{r}
wsp_table[, 2]
```

```{r, eval = F}
 wsp_table[, 3] #what's wrong?
```

```{r}
wsp_table[c(2:5), 2]
```

### Dataframe queries

#### Function *tapply* / table apply 

**What is the average wingspan PER insect species?**

```{r}
mean(x = wsp_table$wingspan[wsp_table$insect == 'dragonfly']) # == !
#- more elegant: function 'tapply'
tapply(X = wsp_table$wingspan, INDEX = wsp_table$insect, FUN = mean)
```

#### Subsetting 

##### simple: function *subset*

```{r}
wsp_sub <- subset(x = wsp_table, subset = wsp_table$wingspan <= 65)
nrow(wsp_sub) # try also: dim(wsp_sub),dim(wsp_sub)[1]
wsp_sub
```

##### A bit more complicated: function *which*

```{r}
wsp_tmp <- which(x = wsp_table$wingspan <= 65)
wsp_tmp
wsp_which <- wsp_table[wsp_tmp,]
wsp_which
```

### Data Manipulation
#### Removing rows/columns 

Omit rows using a minus sign

```{r}
x <- wsp_which
x
x2 <- x[-c(1, 3), ] #remove rows
x2
```


Omit columns

```{r}
x22 <- x$wingspan #removes (actually keeps) columns
x22
```

#### Adding rows/columns I

##### Combine data with *rbind()*

```{r}
x3 <- rbind(x2, wsp_table)
x3[1:10,] # n.b. new row names 41, 91 !
```

##### Combine data with *cbind()*

```{r}
x4 <- cbind(vector_of_numbers[1:3], vector_of_numbers[3:5])
x4
colnames(x4) <- c('col1', 'col2')
rownames(x4) <- c('A', 'B', 'C')
x4
```

## List

#### list per grouping factor (here: insect type)

```{r}
wsp_list <- list(dragonfly = wsp_drfly2, butterfly = wsp_butfly2)
str(wsp_list)
wsp_list$dragonfly # == wsp_list[['dragonfly']]
lapply(X = wsp_list, FUN = mean) # OR: mean(wsp_list$dragonfly)
```

## Matrix 

Definition per row or col, looses headers!
matrix setup per row

```{r}
wsp_matrix <- matrix(data = c(wsp_drfly2, wsp_butfly2), nrow = 2)
wsp_matrix
# Does this format ring a bell????
rownames(x = wsp_matrix) <- c('dragonfly', 'butterfly')
wsp_matrix[, 1]
wsp_matrix['dragonfly', ] # == wsp_matrix[1, ]
```

matrix setup per column

```{r}
wsp_matrix <- matrix(data = c(wsp_drfly2, wsp_butfly2), ncol = 2)
wsp_matrix
#head(wsp_matrix)
#str(wsp_matrix)
wsp_matrix[, 1]
colnames(x = wsp_matrix) <- c('dragonfly', 'butterfly')
#wsp_matrix$dragonfly
wsp_matrix[, 'dragonfly'] # == wsp_matrix[, 1]
```

#### Extracting data from matrix 

```{r}
#wsp_matrix <- matrix(c(wsp_drfly2,wsp_butfly2), ncol=2)
#colnames(wsp_matrix) <- c('dragonfly','butterfly')

mean(x = wsp_matrix[, 1])
apply(X = wsp_matrix, MARGIN = 2, FUN = mean)
#check command, what is meant with 'by 2'?
#check command, what happens if you use 1 instead?
```

## Missing Data NA

```{r}
wsp_fly2 <- c(1,5,4,11,8,2,NA,4,2,9,8,NA)
mean(x = wsp_fly2)
# type ?mean  # How to deal with missing data?

mean(x = wsp_fly2, na.rm=T)
```

```{r}
wsp_l <- list(dragonfly = wsp_drfly2, butterfly = wsp_butfly2, fly = wsp_fly2)
lapply(X = wsp_l, FUN = mean)
lapply(X = wsp_l, FUN = mean, na.rm = T)
```

## Data Manipulation: Summary

```{r}
#TRY
# class(wsp_table) #class(wsp_matrix) #class(wsp_list)
# on vectors: #class(insect_type) # class(wsp_l$dragonfly)
# on skalars: #class(what_is_one_plus_one)
```

<br><br>

# Managing Data 

## Saving and Reading Files 

```{r}
getwd() # to change use setwd(), e.g. setwd('../Output')
write.csv(x = wsp_table, file = "wsp.csv", row.names = FALSE)
wsp <- read.csv(file = "wsp.csv", header = T)
head(x = wsp)
```

<br><br>

# Plotting with R 

## Plotting: Basics

**plot()** is the universal function for scatter plots or line diagrams

```{r}
plot(x = wsp)
plot(x = wsp$wingspan)
```


Use command *ifelse* to define colors PER species

```{r}
color_for_insect_type <- ifelse(test = wsp$insect == 'dragonfly', yes = 'red', no = 'blue')
# color_for_insect_type
plot(x = wsp$wingspan, col= color_for_insect_type)
```


**par()** to pimp up your plot

```{r}
?par #to set graphical parameters
 plot(x = wsp$wingspan, col = color_for_insect_type, pch = 3,
      xlab = "unsorted values", ylab = "wsp [mm]", cex = 3,
      cex.lab = 1.2, cex.axis = 0.5, las = 2,
      col.axis = "blue")
```

## Plotting: Practical thingies 
### plot-function
#### Abline - Drawing lines (h=horizontal, v= vertical, or "intercept, slope")

```{r}
?abline
plot(x = wsp$wingspan, col = color_for_insect_type)
abline(v = 12.5, lty = "dashed", col = "red")
                             #lty = line type
abline(h = 75, lty = "dotted")
```


#### Drawing *arrows* and adding *text*

```{r}
plot(x = wsp$wingspan, col = color_for_insect_type,pch = 16, cex = 3)
arrows(x0 = 5, y0 = 85, x1 = 5, y1 = 82, col = "purple")
arrows(x0 = 15, y0 =  70, x1 = 17, y1 =  75, col = "black", lwd = 3) #lwd=line width
text(x = 3, y = 70, labels = "DRAGONFLY",cex=0.8)
```

## Other plot types

```{r}
#First, create new data set called " mydata" with new column "wind_speed":
wind_speed <- rep(x = 50:70, length.out = nrow(wsp))
mydata <- cbind(wsp, wind_speed)
head(x = mydata)
```

### *boxplot*, histogram (*hist*)

```{r}
par(mfrow = c(1, 2))
boxplot(formula = mydata$wingspan ~ mydata$wind_speed)
hist(x = mydata$wingspan, main = '')
```

### *hist* continued - the role of breaks and freq = F

```{r}
par(mfrow = c(1,2))
hist(x = mydata$wingspan, breaks = 2, main = '')
hist(x = mydata$wingspan, 
     breaks = seq(min(mydata$wingspan),
                  max(mydata$wingspan), 
                  by = 3), 
     freq = F, col = 'lightgrey', main = '')
```


```{r}
my_h <- hist(x = mydata$wingspan, breaks = seq(min(mydata$wingspan), max(mydata$wingspan), by = 3), plot = FALSE)
my_h
```


### barplot

```{r}
my_bp <- table(mydata$wingspan)
my_bp # note the difference to histograms ($counts)!
barplot(my_bp) # WHY is there a maximum of 3 counts in the histogram and 2 in the barplot?
```


### Conditional plot *coplot*

```{r}
coplot(formula = mydata$wingspan ~ mydata$wind_speed | mydata$insect, pch = 16)
```

## Plotting: Margins and multiple figures

### Graphics device division

```{r}
par(mfrow = c(1, 2), mar = c(5, 4, 1, 1), oma = c(1.5, 2, 1, 1))
plot(x = mydata$wingspan, col = color_for_insect_type, pch = 16, cex = 3)
boxplot(formula = mydata$wingspan ~ mydata$insect)
box(which = 'plot', col = 'red')
box(which = 'figure', col = 'blue', lty = 'dashed')
box(which = 'outer', col = 'green' , lwd = 5)
```

## Saving Plots

works for jpg, pdf,png, wmf and postscript

```{r}
#- Single Plot
pdf(file = 'MyPdf.pdf', width =  10, height = 10)
with(data = mydata, expr = {coplot(wingspan ~ wind_speed | insect, pch=16)})
dev.off()
#- Multiple Plot
jpeg(file = 'MyJPG.jpg')
par(mfrow = c(1, 2))
barplot(my_bp, col = rainbow(17), main = 'Barplot', xlab = 'wingspan', ylab = 'Count')
hist(x = mydata$wingspan, 
     breaks = seq(min(mydata$wingspan), 
                  max(mydata$wingspan), 
                  by = 3), 
     freq = T, 
     col = rainbow(6))
dev.off()
#n.b.! Correct output folder path!
```

<br><br>

# Programming with R (shortly) 

## Simulate your own data

```{r}
seq_data  <- seq(from = 1, to = 100, by = 1)
uni_data  <- runif(n = 100, min = 1, max = 10)
norm_data <- rnorm(n = 100, mean = 5, sd = 0.5)
#simulate trend:
trend_data <- 0.1 * seq_data + rnorm(n = length(seq_data), mean = 0, sd = 1)

plot(x = seq_data, y = uni_data, pch = 15)
points(x = norm_data, pch = 20, col = 'red')
points(x = trend_data, pch = 17, col = 'blue')
```

### Just for fun: some stats....

```{r}
my.lm <- lm(formula = trend_data ~ seq_data)
#abline(my.lm)
summary(object = my.lm) #surprise....
```

## The *for* loop

```{r}
#- for repetitive operations
#- *i* is used as variable that is incremented
for(i in 1:4)
{ #begin
  print(x = i)
  if (i == 4) print (x = 'end of lop')
} # end
?'for'
#other useful control flow constructs are: *while*, *repeat*, *if*
```

## Writing your own function

```{r}
my_fct <- function(a) {-8.13 + (12.55 * a)}
my_prob <- function(x) {1/ (1 + exp(x = -x))}

# ff = e.g. frequency of forest in a cell
ff <- seq(from = 0, to = 1 ,by = 0.05)

logit_prob <- my_fct(a = ff)
pres_prob <- my_prob(x = logit_prob)
#pres_prob
plot(x = ff,y = pres_prob, type = 'l')

plot(logit_prob ~ff)
plot(pres_prob~logit_prob)
```


### logistic regression and inverse logit

```{r}
seq_data  <- seq(from = 0, to = 1, by=0.05)
#simulate trend:
#tmp <- pbinom(1:length(seq_data),length(seq_data),0.7)
#trend_data <-  tmp +  rnorm(length(tmp),mean=0,sd=0.01)
trend_data <- pbinom(q = 1:length(seq_data), size = length(seq_data), prob = 0.7)

plot(x = seq_data, y = trend_data, pch = 15)
#plot(x=seq_data,y=tmp, pch=15)

my.lm <- glm(formula = trend_data ~ seq_data, family = 'binomial', na.action = na.omit)
myintercept <- my.lm$coefficients[[1]]
myslope <-  my.lm$coefficients[[2]]
my_fct <- function(a, intcpt, slp) {intcpt + (slp * a)}
my_prob <- function(x) {1 / (1 + exp(x = -x))}

# ff = e.g. frequency of forest in a cell
ff <- seq(from = 0, to = 1 ,by = 0.05)

logit_prob <- my_fct(a = ff, intcpt = myintercept, slp = myslope)
pres_prob <- my_prob(x = logit_prob)
#pres_prob
plot(x = ff, y = pres_prob, type = 'l')
```

<br>
# END



